The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2011-07-21 projects/slrsim/paper/slrsim.tex
Paper final draft.

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper]{geometry}
\usepackage{color}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners led to many false positive results across a wide
range of divergence levels, while others resulted in very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. The performance of the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. \textcolor{red}{As no combination
  of aligner plus filter significantly outperformed the best aligner
  alone,} we conclude that for the sitewise evolutionary analysis of
most vertebrate, insect and fungal phylogenies, the use of unfiltered
PRANK alignments is recommended.

@@

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA} . This suggests that well-folded protein
regions will experience fewer biological indels---and will therefore
be less susceptible to alignment error---than less structured
regions.

@@

A limitation of the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
used their expected general effect of alignment error (an increase in
false positives due to misalignment of non-homologous codons) as a
proxy by which to compare different methods, allowing for the
conclusion that PRANK was the least error-prone aligner in their
analysis. However, the absolute number of false positives remained
uncertain and there was the possibility of conflating multiple sources
of error: in addition to alignment error, the authors noted that gene
mis-annotation was responsible for many apparent false positives, and
there is also an expected error rate from the likelihood inference
method itself. This limitation leaves important and interesting
questions, regarding the nature of alignment error and its
quantitative impact on the detection of positive selection unanswered
by empirical studies.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. \textcolor{red}{GUIDANCE, which measures
  the robustness of alignment regions to perturbations in the guide
  tree used for progressive alignment, has also been proposed as an
  alignment confidence score
  \citep{Penn2010Alignment}}. Unfortunately, despite their widespread
use, the impact of the many available \textcolor{red}{alignment
  scoring and filtering} methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

@@

The Indelible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons and $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4.
Indel lengths were drawn from a discretized
power-law distribution with an exponential decay parameter of 1.8 and
a maximum value of 40, yielding a mean indel length of 3.33 codons and
standard deviation of 5.51 codons. This power-law model of indel
lengths is well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and manual
inspection of alignments from a range of parameter values identified
the chosen model parameters as resulting in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was varied between 0 and 0.2 indel events per substitution
per site.

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. Indelible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were scaled before
simulation to correct for the difference between our definition of
branch lengths as the number of synonymous substitutions per
synonymous site ($dS$) and Indelible's interpretation of branch length
as the average number of substitutions per codon ($t$)
\citep{Fletcher2010Effect}. They are related approximately by
$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
the mean $\omg$ across all sites. $S$ is approximately 0.3 when
$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
for our chosen distribution is 0.277, yielding a $dS$-to-$t$
conversion factor of 1.48 for all simulations performed.

@@

\textcolor{red}{In contrast to Gblocks, GUIDANCE and T-Coffee produce
  individual scores for each residue,} \textcolor{red}{ allowing for
  individual residues to be masked instead of entire columns. GUIDANCE
  generates many replicate alignments, each using a slightly perturbed
  guide tree, and assigns to each residue a score from 0 to 1 based on
  how consistently it was placed in the replicate alignments. We ran
  GUIDANCE with 100 replicates using MAFFT as the aligner, except when
  \prankc was the original aligner, in which case 30 replicate
  alignments were produced using \pranka. T-Coffee calculates the
  consistency between an input multiple alignment and independently
  calculated pairwise alignments \citep{Notredame2003Using}, rounding
  and normalizing residue scores into integers between 0 and
  9. T-Coffee was run using its default settings and the {\em
    evaluate\_mode --output=score\_ascii} command-line parameters to
  output alignment scores.}

\textcolor{red}{To filter alignments based on these residue-wise
  scores, a cutoff threshold was chosen for each method (0.5 for
  GUIDANCE and 5 for T-Coffee) and residues equal to or below that
  threshold were masked. On a per-alignment basis, if the default
  threshold caused greater than 50\% of residues to be masked, then
  the threshold was relaxed to the highest value for which at least
  50\% of residues remained unmasked. We found this adjustment
  necessary because the scores from GUIDANCE and T-Coffee were
  strongly affected by the simulation conditions, with much lower
  average scores at higher indel rates and divergences. Requiring at
  least 50\% of residues to remain unmasked ensured that enough data
  were available for meaningful evolutionary analysis, mimicking
  typical treatment of real data sets.}

Two unrealistic but informative datasets were produced to serve as controls. First,
the true simulated alignment was included in order to evaluate the \sw
performance without any alignment error. Second, an additional
filtering method was constructed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue reflecting how correctly
is has been placed in the inferred alignment. The approach taken was
to calculate, for each residue, the branch length of the correct
sub-tree (defined as the sub-tree connecting all sequences to which
the current residue was correctly aligned) divided by the branch
length of the total aligned sub-tree (defined as the sub-tree
connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
reflects the expectation that correctly-aligned evolutionary branch
length is the main source of information from which \sw inference
methods derive their power. We refer to this method as the 'optimal'
filtering method. Scores were handled in a manner similar to GUIDANCE and
T-Coffee, using a score threshold of 0.5.

@@

In order to compare \sw estimates from different alignments, a single
sequence from each tree was chosen as the reference (arrows, Figure
\ref{fig_1}) and all \sw statistics were mapped from alignment columns
to sequence positions in the reference sequence. This approach bears
some resemblance to the process of mapping alignment-based
evolutionary estimates onto a single member of the alignment for
further analysis and integration with other genome-referenced data (as
is often done, for example, using mammalian alignments and a human
reference). As a result of this reference sequence based mapping,
sites which were deleted in the reference sequence or inserted in a
lineage not ancestral to the reference were \textcolor{red}{not
  included in the final performance analysis.}

@@

When the indel rate was greater than zero, performance levels varied
significantly for different tree sizes, alignment algorithms, and
evolutionary divergences. Figure \ref{fig_3} shows the same
performance measurements as Figure \ref{fig_2} for simulations without
indels (gray lines, Figure \ref{fig_3}) and with indels (black lines,
Figure \ref{fig_3}) analyzed using three different aligners and the
true alignment. For the indel simulations, the indel rate here was held
constant at 0.1 indel event per substitution.
\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.75]{fig3.pdf}
\includegraphics[scale=1.25]{fig3.pdf}
\end{center}
\caption{Sequences were simulated without indels (gray lines) or with
  indels (black lines) using one of three tree shapes, aligned with
  one of three aligners, and analyzed with SLR; true alignments were
  separately analyzed with SLR (solid black lines). One hundred
  replicate alignments were simulated for each data point. The
  performance of each dataset, as measured by four summary statistics,
  is plotted as a function of the mean path length (MPL)
  divergence. From top to bottom: true positive rate (TPR) at the
  recommended cutoff threshold; false positive rate (FPR) at the
  recommended cutoff threshold; TPR at a 1\% FPR threshold; Pearson's
  correlation coefficient between the true and inferred $\omg$.}
\label{fig_3}
\end{figure}

@@

Looking at the TPR results for inferred alignments, we observed that
in the 6-taxon tree the three aligners formed a cluster of lines well
below the true alignment value, indicating similar tendencies among
the different aligners to produce false negatives in the smaller
tree. In larger trees the different aligners showed a wider spread of
TPR values, but even \prankc{} showed a 5--10\% reduction compared to
the true alignment at MPL$=$1.0. These results show that the
introduction of false negatives is a significant and seemingly
unavoidable result of alignment error at medium to high divergence
levels (MPL $>$ 0.5), with even the most successful aligner producing
a marked reduction in TPR compared to the true
alignment. \textcolor{red}{The \tpr and $\omg$ results in the larger
  two trees were qualitatively similar to the TPR results, showing
  that the aligners tested led to different levels of sitewise
  performance even when controlling for actual error rates or
  assessing the sitewise $\omg$ correlation.}

@@

We found it useful to combine divergence estimates from Table
\ref{table_2} with the results from Figure \ref{fig_3} to characterize
the combined effects of alignment error at different commonly analyzed
levels of divergence. For example, at a human-mouse divergence level
(MPL$=$0.2) misalignment had little impact on the TPR regardless of
what aligner was used. However, ClustalW yielded a notably higher FPR
than MAFFT or \prankc, and the error-controlled \tpr{} was
correspondingly lower for ClustalW in all three trees. Thus, at low
divergences we found that false positives were the main source of
error from misalignment, and different aligners had highly variable
tendencies to produce false positive results. At higher vertebrate and
\Dr divergence levels (MPL$=$0.8--1.0) false negatives became much
more prevalent. The TPR for all inferred alignments was virtually zero
in the 6-taxon tree, underscoring the necessity of including many
species in the analysis of highly diverged sequences. In the two
larger trees, \prankc resulted in very few additional false positives,
but it suffered a 5--10\% reduction in TPR relative to the true
alignment. Meanwhile, ClustalW showed a 50\% TPR reduction and
maintained a strongly elevated FPR. At higher divergences and in
larger trees, false negatives were thus the most persistent effect of
alignment error, causing a marked reduction in \sw power even with the
best-performing aligner. \textcolor{red}{Overall, the ranking of
  aligners was clear, with \prankc performing better than MAFFT, and
  MAFFT performing better than ClustalW, across all performance
  measures and MPL divergence levels.}

\begin{figure*}[t]
\begin{center}
%\includegraphics[scale=0.8]{fig4.pdf}
\includegraphics[scale=0.8]{fig4.pdf}
\end{center}
\caption{Sequences were simulated with indels using one of three tree
  shapes and a range of indel rates and mean path length (MPL)
  divergence levels. Alignments were inferred with one of four
  aligners and analyzed with SLR; true alignments were separately
  analyzed with SLR. One hundred replicates were simulated for each
  set of conditions. Each square is colored according to the
  performance at a given (indel rate, MPL) pair as measured by one of
  three summary statistics: (a) the true positive rate (TPR) at the
  recommended cutoff threshold, (b) the false positive rate (FPR) at
  the recommended cutoff threshold, or (c) the TPR at a 1\% FPR
  threshold.}
\label{fig_4}
\end{figure*}

@@

To explore the effects of alignment error across a wider range of
simulation conditions, we extended the simulations of Figure
\ref{fig_3} across multiple indel rates. Figure \ref{fig_4} shows
heatmaps of the TPR and FPR for ClustalW, \prankc and the true
alignment (Figure \ref{fig_4}A,B) and a heatmap of the
error-controlled \tpr for all aligners tested (Figure
\ref{fig_4}C). The results from Figure \ref{fig_3}, which were
simulated with an indel rate of 0.1, correspond to the middle row of
each panel in Figure \ref{fig_4}; rows above and below the middle row
represent higher and lower indel rates, respectively. \textcolor{red}{Similarly, the
bottom row of each panel in Figure \ref{fig_4} was simulated with an
indel rate of 0 and corresponds to the 'No Indels' data in Figure
\ref{fig_3}.}

@@

\prankc{} and ClustalW both showed a qualitatively similar pattern of
elevated FPRs in the 6-taxon tree (Figure \ref{fig_4}B), but their
behavior diverged significantly in the 17-taxon and 44-taxon trees. In
the 17-taxon tree, \prankc{} only showed an elevated FPR compared to
the true alignment at very high indel rates and divergence levels, but
the ClustalW FPR increased steadily with the indel rate, quadrupling
in value from the lowest to highest indel rate. Interestingly, for any
given indel rate, the ClustalW FPR showed little variation across the
range of divergence levels. This result was counter-intuitive, as we
expected alignment errors to become more common as divergence
increased and the number of observed indel events grew. Furthermore,
\prankc{} behaved according to our expectations, showing increased
FPRs only at the highest divergences and indel rates in the 17-taxon
tree. The FPR results in the 44-taxon tree confirmed the strange
effect of ClustalW’s alignments on the \sw FPR: at the highest indel
rates, ClustalW showed a negative relationship between FPR and
divergence---exactly opposite to the trend we expected. \prankc{}'s
FPR in the 44-taxon tree was equal to or below that of the true
alignment under almost all conditions.

The error-controlled \tpr results (Figure \ref{fig_4}C) provide a
comprehensive picture of the effect of alignment error on the detection of
\sw positive selection. The two aligners not shown in the two other
panels, MAFFT and \pranka, exhibited \tpr values intermediate to those
from ClustalW and \prankc across the range of parameters tested, with \pranka performing
better than MAFFT. As
expected, performance was very similar between aligners at very low
indel rates. At higher indel rates, most aligners yielded similar
patterns of low \tpr in the 6-taxon tree, but in the larger two trees
ClustalW and MAFFT alignments were unable to achieve high \tpr
values, presumably due largely to their elevated FPR in those trees.

@@

Overall, our alignment filtering simulations found that filtering
methods based on GUIDANCE and T-Coffee scores had the ability to mask
away many residues causing false positives and false negatives from
ClustalW alignments, with high levels of improvement across a wide
range of divergence levels and indel rates. Filtering was less
beneficial when applied to \prankc alignments, with T-Coffee reducing
performance and GUIDANCE yielding only mild \tpr
improvements. Importantly, GUIDANCE only showed improved performance
at high divergence levels (e.g., MPL$>$1.6), well above those found in
commonly-analyzed groups of species. Thus, the use of unfiltered
\prankc alignments would yield the best performance for detecting \sw
positive selection when analyzing protein-coding sequences across a
wide range of indel rates and commonly encountered sequence divergence
levels.

>>>>

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}

@@

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners led to many false positive results across a wide
range of divergence levels, while others resulted in very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. Controls using the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. As no combination of aligner plus
filter significantly outperformed the best aligner alone, we conclude
that for the sitewise evolutionary analysis of most vertebrate, insect
and fungal phylogenies, the use of unfiltered PRANK alignments is
recommended.

@@

The protein structure and the evolutionary divergence of a dataset may
also contribute to the effects of alignment error. Differently
structured protein regions show variable tolerance to biological
indels, with indels more common in extracellular and transmembrane
proteins than in highly folded enzymes and housekeeping genes
\citep{delaChaux2007DNA}. This suggests that well-folded protein
regions will experience fewer biological indels---and will therefore
be less susceptible to alignment error---than less structured
regions.

@@

A limitation of the analysis of error in empirical datasets is the
lack of a benchmark set of true alignments and positively-selected
sites. Markova-Raina and Petrov \citeyearpar{Markova-Raina2011High}
used their expected general effect of alignment error (an increase in
false positives due to misalignment of non-homologous codons) as a
proxy by which to compare different methods, allowing for the
conclusion that PRANK was the least error-prone aligner in their
analysis. However, the absolute number of false positives remained
uncertain and there was the possibility of conflating multiple sources
of error: in addition to alignment error, the authors noted that gene
mis-annotation was responsible for many apparent false positives, and
there is also an expected error rate from the likelihood inference
method itself. This limitation leaves important and interesting
questions, regarding the nature of alignment error and its
quantitative impact on the detection of positive selection, unanswered
by empirical studies.

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. GUIDANCE, which measures the robustness of
alignment regions to perturbations in the guide tree used for
progressive alignment, has also been proposed as an alignment
confidence score \citep{Penn2010Alignment}. Unfortunately, despite
their widespread use, the impact of the many available alignment
scoring and filtering methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

@@

The INDELible program \citep{Fletcher2009INDELible} was used to
simulate codon sequences with indels along each phylogenetic tree. The
length of the root sequence was set to 500 codons and $\bm{\kappa}$ (the
ratio of transition to transversion substitutions) was fixed at 4.
Indel lengths were drawn from a discretized
power-law distribution with an exponential decay parameter of 1.8 and
a maximum value of 40, yielding a mean indel length of 3.33 codons and
standard deviation of 5.51 codons. The power-law model of indel
lengths is well-supported by empirical studies
\citep{Benner1993Empirical,Cartwright2009Problems} and manual
inspection of alignments from a range of parameter values identified
the chosen model parameters as resulting in alignments most closely
resembling those encountered in vertebrate alignments. The ratio of
insertion to deletion events was set to 1, and the rate of indel
formation was varied between 0 and 0.2 indel events per substitution
per site.

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

Branch lengths for each of the simulation trees were scaled before
simulation to correct for the difference between our definition of
branch lengths as the number of synonymous substitutions per
synonymous site ($dS$) and INDELible's interpretation of branch length
as the average number of substitutions per codon ($t$)
\citep{Fletcher2010Effect}. They are related approximately by
$t=3(NdN+SdS) = 3dS(\bar{\omg{}}N+S)$, where $N$ and $S$ are the
proportion of nonsynonymous and synonymous sites and $\bar{\omg}$ is
the mean $\omg$ across all sites. $S$ is approximately 0.3 when
$\bm{\kappa}=4$ \citep{Yang1998Synonymous} and the mean $\omg$ ratio
for our chosen distribution is 0.277, yielding a $dS$-to-$t$
conversion factor of 1.48 for all simulations performed.

@@

GUIDANCE and T-Coffee produce individual scores for each residue,
allowing individual residues to be masked instead of entire
columns. GUIDANCE generates many replicate alignments, each using a
slightly perturbed guide tree, with either MAFFT or \pranka as the
bootstrap aligner. The program then assigns to each residue from the
input alignment a score from 0 to 1 based on how consistently it was
placed in the replicate alignments. In order to maximize the
similarity between the input aligner and the bootstrap aligner, we ran
GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
and with 30 \pranka replicates when filtering \prankc
alignments. T-Coffee calculates the residue-wise consistency between
an input multiple alignment and independently calculated pairwise
alignments \citep{Notredame2003Using}, rounding and normalizing
residue scores into integers between 0 and 9. T-Coffee was run using
its default settings and the {\em evaluate\_mode
  --output=score\_ascii} command-line parameters to output alignment
scores.

To filter alignments based on these residue-wise scores, a cutoff
threshold was chosen for each method (0.5 for GUIDANCE and 5 for
T-Coffee) and residues equal to or below that threshold were
masked. On a per-alignment basis, if the default threshold caused
greater than 50\% of residues to be masked, then the threshold was
relaxed to the highest value for which at least 50\% of residues
remained. We found this adjustment necessary because the scores from
GUIDANCE and T-Coffee were strongly affected by the simulation
conditions, with much lower average scores at higher indel rates and
divergences. Requiring at least 50\% of residues to remain unmasked
ensured that enough data were available for meaningful evolutionary
analysis, mimicking typical treatment of real data sets.

Two unrealistic but informative datasets were produced to serve as controls. First,
the true simulated alignment was included in order to evaluate the \sw
performance without any alignment error. Second, an additional
filtering method was constructed to represent an unattainable
best-case scenario for sequence filtering, using knowledge of the true
alignment to assign a score to each residue reflecting how correctly
is has been placed in the inferred alignment. The approach taken was
to calculate, for each residue, the branch length of the correct
sub-tree (defined as the sub-tree connecting all sequences to which
the current residue was correctly aligned) divided by the branch
length of the total aligned sub-tree (defined as the sub-tree
connecting all sequences with non-gap residues at the current
alignment column). This residue-wise score ranges from 0 to 1 and
reflects the expectation that correctly-aligned evolutionary branch
length is the main source of information from which \sw inference
methods derive their power. We refer to this method as the `optimal'
filtering method. Scores were handled in a manner similar to GUIDANCE and
T-Coffee, using a score threshold of 0.5.

@@

In order to compare \sw estimates from different alignments, a single
sequence from each tree was chosen as the reference (arrows, Figure
\ref{fig_1}) and all \sw statistics were mapped from alignment columns
to sequence positions in the reference sequence. This approach
corresponds to the process of mapping alignment-based evolutionary
estimates onto a single member of the alignment for further analysis
and integration with other genome-referenced data (as is often done,
for example, using mammalian alignments and a human reference). As a
result of this reference sequence based mapping, sites which were
deleted in the reference sequence or inserted in a lineage not
ancestral to the reference were not included in the final performance
analysis.

@@

When the indel rate was greater than zero, performance levels varied
significantly for different tree sizes, alignment algorithms, and
evolutionary divergences. Figure \ref{fig_3} shows the same
performance measurements as Figure \ref{fig_2} for simulations without
indels (gray lines, Figure \ref{fig_3}) and with indels (black and
textured lines, Figure \ref{fig_3}) analyzed using three different
aligners (ClustalW, MAFFT, and \prankc) and the true alignment. (We
omitted results for \pranka alignments from Figure \ref{fig_3} in
order to reduce visual clutter; a full comparison of all aligners
tested is shown in Figure \ref{fig_4}C and discussed in the next
section.) For the indel simulations, the indel rate here was held
constant at 0.1 indel event per substitution.
\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.75]{fig3.pdf}
\includegraphics[scale=1.25]{fig3.pdf}
\end{center}
\caption{Sequences were simulated without indels (solid gray lines) or with
  indels (solid black and textured lines) using one of three tree shapes, aligned with
  one of three aligners, and analyzed with SLR; true alignments were
  separately analyzed with SLR (solid black lines). One hundred
  replicate alignments were simulated for each data point. The
  performance of each dataset, as measured by four summary statistics,
  is plotted as a function of the mean path length (MPL)
  divergence. From top to bottom: true positive rate (TPR) at the
  recommended cutoff threshold; false positive rate (FPR) at the
  recommended cutoff threshold; TPR at a 1\% FPR threshold; Pearson's
  correlation coefficient between the true and inferred $\omg$.}
\label{fig_3}
\end{figure}

@@

Looking at the TPR results for inferred alignments, we observed that
in the 6-taxon tree the three aligners formed a cluster of lines well
below the true alignment value, indicating similar tendencies among
the different aligners to produce false negatives in the smaller
tree. In larger trees the different aligners showed a wider spread of
TPR values, but even \prankc{} showed a 5--10\% reduction compared to
the true alignment at MPL$=$1.0. These results show that the
introduction of false negatives is a significant and seemingly
unavoidable result of alignment error at medium to high divergence
levels (MPL $>$ 0.5), with even the most successful aligner producing
a marked reduction in TPR compared to the true alignment. The \tpr and
$\omg$ results in the larger two trees were qualitatively similar to
the TPR results, showing that the aligners tested led to different
levels of sitewise performance even when controlling for actual error
rates or assessing the sitewise $\omg$ correlation.

@@

We found it useful to combine divergence estimates from Table
\ref{table_2} with the results from Figure \ref{fig_3} to characterize
the combined effects of alignment error at different commonly analyzed
levels of divergence. For example, at a human-mouse divergence level
(MPL$=$0.2) misalignment had little impact on the TPR regardless of
what aligner was used. However, ClustalW yielded a notably higher FPR
than MAFFT or \prankc, and the error-controlled \tpr{} was
correspondingly lower for ClustalW in all three trees. Thus, at low
divergences we found that false positives were the main source of
error from misalignment, and different aligners had highly variable
tendencies to produce false positive results. At higher vertebrate and
\Dr divergence levels (MPL$=$0.8--1.0) false negatives became much
more prevalent. The TPR for all inferred alignments was virtually zero
in the 6-taxon tree, underscoring the necessity of including many
species in the analysis of highly diverged sequences. In the two
larger trees, \prankc resulted in very few additional false positives,
but it suffered a 5--10\% reduction in TPR relative to the true
alignment. Meanwhile, ClustalW showed a 50\% TPR reduction and
maintained a strongly elevated FPR. At higher divergences and in
larger trees, false negatives were thus the most persistent effect of
alignment error, causing a marked reduction in \sw power even with the
best-performing aligner. Overall, the ranking of aligners was clear,
with \prankc performing better than MAFFT and MAFFT performing better
than ClustalW across all performance measures and MPL divergence
levels.

\begin{figure*}[t]
\begin{center}
%\includegraphics[scale=0.8]{fig4.pdf}
\includegraphics[scale=0.8]{fig4.pdf}
\end{center}
\caption{Sequences were simulated with indels using one of three tree
  shapes (t-, 17- or 44-taxon) and a range of indel rates and mean
  path length (MPL) divergence levels. Alignments were inferred with
  one of four aligners (ClustalW, MAFFT, \pranka{}, \prankc) and
  analyzed with SLR; true alignments were separately analyzed with
  SLR. One hundred replicates were simulated for each set of
  conditions. Each cell is colored according to the performance at a
  given (indel rate, MPL) pair as measured by one of three summary
  statistics: (a) the true positive rate (TPR) at the recommended
  cutoff threshold, (b) the false positive rate (FPR) at the
  recommended cutoff threshold, or (c) the TPR at a 1\% FPR
  threshold. Results for MAFFT and \pranka are omitted from (a) and
  (b); as in (c) they show characteristics intermediate between
  ClustalW and \prankc.}
\label{fig_4}
\end{figure*}

@@

To explore the effects of alignment error across a wider range of
simulation conditions, we extended the simulations of Figure
\ref{fig_3} across multiple indel rates. Figure \ref{fig_4} shows
heatmaps of the TPR and FPR for ClustalW, \prankc and the true
alignment (Figure \ref{fig_4}A,B) and a heatmap of the
error-controlled \tpr for all aligners tested (Figure
\ref{fig_4}C). (MAFFT and \pranka were omitted from Figure
\ref{fig_4}A--B to save space, but their performance fell between that
of ClustalW and \prankc for all performance measures, with \pranka
slightly outperforming MAFFT; a comprehensive set of TPR, FPR, and
\tpr results can be found in Supplementary Figure S1.)
The results from Figure \ref{fig_3}, which were simulated with an
indel rate of 0.1, correspond to the middle row of each panel in
Figure \ref{fig_4}; rows above and below the middle row represent
higher and lower indel rates, respectively. Similarly, the bottom row
of each panel in Figure \ref{fig_4} was simulated with an indel rate
of zero and corresponds to the `No Indels' data in Figure \ref{fig_3}.

@@

\prankc{} and ClustalW both showed a qualitatively similar pattern of
elevated FPRs in the 6-taxon tree (Figure \ref{fig_4}B), but their
behavior diverged significantly in the 17-taxon and 44-taxon trees. In
the 17-taxon tree, \prankc{} only showed an elevated FPR compared to
the true alignment at very high indel rates and divergence levels, but
the ClustalW FPR increased steadily with the indel rate, quadrupling
in value from the lowest to highest indel rate. Interestingly, for any
given indel rate, the ClustalW FPR showed little variation across the
range of divergence levels. This result was counter-intuitive, as we
expected alignment errors to become more common as divergence
increased and the number of observed indel events grew. Furthermore,
\prankc{} behaved according to our expectations, showing increased
FPRs only at the highest divergences and indel rates in the 17-taxon
tree. The FPR results in the 44-taxon tree confirmed the strange
effect of ClustalW's alignments on the \sw FPR: at the highest indel
rates, ClustalW showed a negative relationship between FPR and
divergence---exactly opposite to the trend we expected. \prankc{}'s
FPR in the 44-taxon tree was equal to or below that of the true
alignment under almost all conditions.

The error-controlled \tpr results (Figure \ref{fig_4}C) provide a
comprehensive picture of the effect of alignment error on the detection of
\sw positive selection. The two aligners not shown in the two other
panels (MAFFT and \pranka) exhibited \tpr values intermediate to those
from ClustalW and \prankc across the range of parameters tested, with \pranka performing
better than MAFFT. As
expected, performance was very similar between aligners at very low
indel rates. At higher indel rates, most aligners yielded similar
patterns of low \tpr in the 6-taxon tree, but in the larger two trees
ClustalW and MAFFT alignments were unable to achieve high \tpr
values, presumably due largely to their elevated FPR in those trees.

@@

Overall, our alignment filtering simulations found that Gblocks rarely
improves alignments for sitewise detection of positive selection, but
filtering methods based on GUIDANCE and T-Coffee scores have a good
ability to mask out misaligned residues that cause false positives and
false negatives in sitewise inference. This beneficial effect is
highly dependent on simulation conditions and the input aligner. For
ClustalW alignments (which, left unfiltered, led to many false
positives and false negatives) both GUIDANCE and T-Coffee showed good
ability to improve sitewise performance, behaving qualitatively
similarly to the `optimal' filter. In order to compare the TPR, FPR
and \tpr values obtained with filtered ClustalW alignments to those
obtained with unfiltered alignments, we ran ClustalW with GUIDANCE
filtering using all three trees and the same range of indel rates and
MPLs as used for Figure \ref{fig_4}; these results are included in
Supplementary Figure S1. GUIDANCE filtering generally
improved the performance of ClustalW alignments, resulting in sitewise
performance comparable to (but not better than) unfiltered MAFFT
alignments.
